Sketch Recognition using Particle Swarm Algorithms

### 2009-07-26 trunk/smallpaper/SmallPaper.tex
final submitted
final after corrections

\documentclass{article}
\usepackage{spconf}
\usepackage{amsmath}
\usepackage[dvips,pdftex]{graphicx}
\usepackage{subfigure}
\usepackage{named}
\usepackage[dvips,pdftex]{hyperref}
\title{Sketch Recognition using Particle Swarm Algorithms}
%\sthanks{Professor}\sthanks{Associate Professor}\sthanks{Professor}\sthanks{Associate Professor}
\twoauthors
  {Maha El Meseery, Mahmoud Fakhr El Din, Samia Mashali}
	{Signals Processing Group \\ Computers and Systems Department \\
  Electronic Research Institute\\Cairo, Egypt\\
	melmeseery@eri.sci.eg, samia@eri.sci.eg}
  {Magda Fayek, Nevin Darwish}
	{ Computer Engineering Department\\
	Faculty of Engineering \\Cairo University\\Cairo, Egypt\\
	magdafayek@gmail.com,ndarwish@ieee.org}
\begin{document}
\maketitle
\begin{abstract}
Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Users draw strokes using a pen and the system immediately interprets their strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. This paper uses Support Vector Machines (SVM) classifier which further improves the final recognition accuracy.  
\end{abstract}
\section{Introduction}
The field of symbol recognition has gained interest in the last few years. Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketches more effectively than using only papers. Increasing the interaction between computers and users in sketch and CAD systems has been the reason for the emerging of a few advanced sketch recognition systems.Sketch recognition is the process of identifying user strokes into a meanful symbol that can be furthur used.  
%Sketch recognition is divided into three main steps preprocessing, segmentation and symbol recognition. The preprocessing phase captures user input stroke points and collects basic information about the stroke then proceeds to remove noise and compute basic statistical and geometrical information. In segmentation phase, the strokes are divided into a set of simple geometrical primitives or segments. In the third phase, sketch recognition, strokes and segments are clustered to formulate symbols that can be recognized by a classifier system.

A wide variety of techniques were suggested for sketch recognition.  A hybrid algorithm was introduced in \cite{earlyprocess} where they generated different sets of segments based on both curvature and speed dominant points, followed by choosing a segmentation with the least error from a generated hybrid set. Their system is however limited to recognizing only specific simple geometric shapes with a set of low level recognizers. Each low level recognizer is designed to recognize only one geometric shape using spatial and geometric information extracted from the stroke.  Yu \cite{meanshift10} introduced a feature area for each primitive and then computed the segmentation error for different types of primitives based on the computed feature area. His system achieved good accuracy in simple shapes (square, ellipse,...) but did not perform well in more complex shapes. Paulson and Hammond\cite{Paleosketch08} introduced a set of low level recognizers that were reported to achieve 98\%  accuracy but their system, similar to all low level recognizers, only identify a small set of simple shapes. One of the problems in these previous systems is that they only rely on curvature and speed data which are supjected to noise. They are also subjected to over and under segmentation problems due to use of thresholds.

@@

The contribution in the system is mainly in attempting to generate the optimal segmentation using PSO which exhibited superiority with respect to simillar algoirthms in various other applications \cite{PolygonApproximationPSO}. The system attempts to find the optimial decomposition of the input stroke using PSO with the help of curvature and speed information. The use of PSO algorithm helps in eliminating the effects of input noise, problems of over and under segmentation that were in pervious systems. The proposed symbol recognizier uses the generated segmenation to compute the feature vector which is used to identify more complex symbols. This enables the system to be scalable for more  complex shapes.

@@

The pervious algorithm segments stroke into set of line segments but storkes dose not always consist of line only. It is more meanful to use curves when segmenting strokes. Therefor, a \textit{second algorithm \textsl{AlgS2}} is used to decompose strokes to set of lines and circular arcs. \textit{\textsl{AlgS2}} has the same problem formulation but different fitness and error functions are used. An attempt is made to fit each segment into line or circular arc, the error of both circle and line fit estimations are computed for each segment $S_s=\widehat{x_lx_k}$, the approximation with the lower error value is the chosen approximation of this segment $S_s$\cite{CruveDivisionSwarm}. The sum of the approximation error of all segments is the total error of the particle. The total error of the particle is computed by equation \ref{eq:errorSwarm2}
 \begin{equation}
E=\sum\nolimits_{s = 1}^M e(D_s) 
\label{eq:errorSwarm2}
\end{equation}where $M$ is the number of segments in the solution as generated by the swarm algorithm, $D_s$ is the minimum approximation error of curve and line approximations $min(d_c,d_l)$ as computed by \cite{CruveDivisionSwarm}.  The fitness is computed by the equation \ref{eq:fitnessSwarm2} \begin{equation}
\max fitness(P_i ) = \frac{1}{{E \times M^k }}
\label{eq:fitnessSwarm2}
\end{equation} where $E$ is the error and $M$ is the number of segments and $k$ is a parameter tweaked to get minimum number of segments. The larger $k$ is, the more effect the number of segments will have. For our system, $k$ is selected to be $0.5$\cite{CruveDivisionSwarm}.

@@

		\subfigure[Algorithm comparison]{\includegraphics[width=5cm]{images/testAlg.jpg} 	\label{fig:test1}}
		\end{center}
		\end{minipage}
		\hfill
  \subfigure [Symbols Comparison] {\includegraphics[width=9cm]{images/test2smallpaper.jpg} \label{fig:test2}}
	\caption{Experiments results :} a) The symbols set used in the dataset.   b) The recognition rate of different algorithms.  c) The recognition rate of each symbol using different algorithms. 
\end{figure*} 
We performed two experiments to test the system. Firstly, we tested recognition accuracy of shapes in the data set with both algorithms (\textsl{AlgS1} and \textsl{AlgS2}). We also implemented the segmentation algorithm described in \cite{earlyprocess} (\textsl{Alg3}) to compare it with our swarm algorithms. Figure \ref{fig:test1} shows the accuracy achieved by each algorithm. The result shows that (\textsl{AlgS2}) gives better performance than Alg3 and AlgS1. 
% Firstly we tested recognition accuracy of shapes in the data set with both algorithms (\textsl{AlgS1} and \textsl{AlgS2}). Figure \ref{fig:test1} shows the accuracy achieved by each algorithm. The two swarm algorithms were tested with and without the ellipse fitting module. Including the ellipse detection module improves the results with both the \textit{DPSO} algorithms.% The results show that both \textit{DPSO} algorithms achieve better result than other algorithms. The two swarm algorithms were tested with and without the ellipse fitting module. Combining the ellipse detection module improves the performance in the tested dataset.

@@

\section{Conclusion and Future Work}
\label{ConclusionandFutureWork}
This paper presented a new approach to sketch recognition using DPSO. Results show that the \textit{DPSO} in general generates an optimized stroke segmentation which improves the final recognition rate.  The final recognition is based on geomterical,statistical and structural features. The experiments evaluated two \textit{DPSO} different segmentation algorithms (AlgS1 and AlgS2) on simple presentation set dataset. Results show that AlgS2 achieves better segmentation and final accuracy than Alg3 \cite{earlyprocess} on the dataset used. The results proved that the system is easily expandable to more and more symbols as it does not depend on low level primitive recognizers but on high level symbol and geometrical features.  The system achieved an average overall improvement of more than 2\% over Alg3.  
%This paper presented a new approach to sketch recognition using PSO. The system uses both speed and curvature information which help improving the \textit{DPSO}. It is noted that the \textit{DPSO} in general generates an optimized stroke segmentation which improves the final recognition rate.  The tradeoff between accuracy achieved and time complexity must be further investigated to achieve better results. The use of statistical moments and structural features improves the recognition rate. The system was tested on 13 different symbols and achieved an accuracy of 91.5\%. The system does not depend on low level recognizer but rather on a set of high level features. This makes the system easily expandable to more and more symbols.

>>>>

\documentclass{article}
\usepackage{spconf}
\usepackage{amsmath}
\usepackage[dvips,pdftex]{graphicx}
\usepackage{subfigure}
\usepackage{named}
\usepackage[dvips,pdftex]{hyperref}
\title{Sketch Recognition using Particle Swarm Algorithms}
%\sthanks{Professor}\sthanks{Associate Professor}\sthanks{Professor}\sthanks{Associate Professor}
\twoauthors
  {Maha El Meseery, Mahmoud Fakhr El Din, Samia Mashali}
	{Signals Processing Group \\ Computers and Systems Department \\
  Electronic Research Institute\\Cairo, Egypt\\
	melmeseery@eri.sci.eg, samia@eri.sci.eg}
  {Magda Fayek, Nevin Darwish}
	{ Computer Engineering Department\\
	Faculty of Engineering \\Cairo University\\Cairo, Egypt\\
	magdafayek@gmail.com,ndarwish@ieee.org}
\begin{document}
\maketitle
\begin{abstract}
Sketch recognition is defined as the process of identifying symbols that users draw using single or multiple strokes. Users draw strokes using a pen and the system immediately interprets their strokes as objects that can be easily manipulated. This paper uses Particle Swarm Optimization Algorithm (PSO) to divide the strokes the user draws into meaningful geometric primitives. These geometric primitives are grouped to formulate symbols which are further identified. The results show that using PSO improves segmentation results which guide the symbol recognition phase. This paper uses Support Vector Machines (SVM) classifier which further improves the final recognition accuracy.  
\end{abstract}
\section{Introduction}
The field of symbol recognition has gained interest in the last few years. Scientists generally and engineers specifically express thoughts and designs using sketches. Engineers use sketches to exchange designs as a natural method of communication rather than writing or speaking. Design engineers need a powerful computer based system with symbol recognition to help them design, manipulate and store sketches more effectively than using only papers. Increasing the interaction between computers and users in sketch and CAD systems has been the reason for the emerging of a few advanced sketch recognition systems.Sketch recognition is the process of identifying user strokes into a meaningful symbols that can be further used.  
%Sketch recognition is divided into three main steps preprocessing, segmentation and symbol recognition. The preprocessing phase captures user input stroke points and collects basic information about the stroke then proceeds to remove noise and compute basic statistical and geometrical information. In segmentation phase, the strokes are divided into a set of simple geometrical primitives or segments. In the third phase, sketch recognition, strokes and segments are clustered to formulate symbols that can be recognized by a classifier system.

A wide variety of techniques were suggested for sketch recognition.  A hybrid algorithm was introduced in \cite{earlyprocess} where they generated different sets of segments based on both curvature and speed dominant points, followed by choosing a segmentation with the least error from a generated hybrid set. Their system is however limited to recognizing only specific simple geometric shapes with a set of low level recognizers. Each low level recognizer is designed to recognize only one geometric shape using spatial and geometric information extracted from the stroke.  Yu \cite{meanshift10} introduced a feature area for each primitive and then computed the segmentation error for different types of primitives based on the computed feature area. His system achieved good accuracy in simple shapes (square, ellipse,...) but did not perform well in more complex shapes. Paulson and Hammond\cite{Paleosketch08} introduced a set of low level recognizers that were reported to achieve 98\%  accuracy but their system, similar to all low level recognizers, only identify a small set of simple shapes. One of the problems in these previous systems is that they only rely on curvature and speed data which are subjected to noise. They are also subjected to over and under segmentation problems due to use of thresholds.

@@

The contribution in the system is mainly in attempting to generate the optimal segmentation using PSO which exhibited superiority with respect to similar algorithms in various other applications \cite{PolygonApproximationPSO}. The system attempts to find the optimal decomposition of the input stroke using PSO with the help of curvature and speed information. The use of PSO algorithm helps in eliminating the effects of input noise, problems of over and under segmentation that were in pervious systems. The proposed symbol recognizer uses the generated segmentation to compute the feature vector which is used to identify more complex symbols. This enables the system to be scalable for more  complex shapes.

@@

The previous algorithm segments stroke into set of line segments but strokes dose not always consist of line only. It is more meaningful to use curves when segmenting strokes. Therefor, a \textit{second algorithm \textsl{AlgS2}} is used to decompose strokes to set of lines and circular arcs. \textit{\textsl{AlgS2}} has the same problem formulation but different fitness and error functions are used. An attempt is made to fit each segment into line or circular arc, the error of both circle and line fit estimations are computed for each segment $S_s=\widehat{x_lx_k}$, the approximation with the lower error value is the chosen approximation of this segment $S_s$\cite{CruveDivisionSwarm}. The sum of the approximation error of all segments is the total error of the particle. The total error of the particle is computed by equation \ref{eq:errorSwarm2}
 \begin{equation}
E=\sum\nolimits_{s = 1}^M e(D_s) 
\label{eq:errorSwarm2}
\end{equation}where $M$ is the number of segments in the solution as generated by the swarm algorithm, $D_s$ is the minimum approximation error of curve and line approximations $min(d_c,d_l)$ as computed by \cite{CruveDivisionSwarm}.  The fitness is computed by the equation \ref{eq:fitnessSwarm2} \begin{equation}
\max fitness(P_i ) = \frac{1}{{E \times M^k }}
\label{eq:fitnessSwarm2}
\end{equation} where $E$ is the error and $M$ is the number of segments and $k$ is a parameter tweaked to get minimum number of segments. The larger $k$ is, the more effect the number of segments will have. For our system, $k$ is selected to be $0.5$\cite{CruveDivisionSwarm}.

@@

		\subfigure[Algorithm comparison]{\includegraphics[width=5cm]{images/testAlg.jpg} 	\label{fig:test1}}
		\end{center}
		\end{minipage}
		\hfill
  \subfigure [Symbols Comparison] {\includegraphics[width=11cm]{images/symbols.jpg} \label{fig:test2}} %symbols.jpg  test2smallpaper.jpg
	\caption{Experiments results :} a) The symbols set used in the dataset.   b) The recognition rate of different algorithms.  c) The recognition rate of each symbol using different algorithms. 
\end{figure*} 
We performed two experiments to test the system. Firstly, we tested recognition accuracy of shapes in the data set with both algorithms (\textsl{AlgS1} and \textsl{AlgS2}). We also implemented the segmentation algorithm described in \cite{earlyprocess} (\textsl{Alg3}) to compare it with our swarm algorithms. Figure \ref{fig:test1} shows the accuracy achieved by each algorithm. The result shows that (\textsl{AlgS2}) gives better performance than Alg3 and AlgS1. 
% Firstly we tested recognition accuracy of shapes in the data set with both algorithms (\textsl{AlgS1} and \textsl{AlgS2}). Figure \ref{fig:test1} shows the accuracy achieved by each algorithm. The two swarm algorithms were tested with and without the ellipse fitting module. Including the ellipse detection module improves the results with both the \textit{DPSO} algorithms.% The results show that both \textit{DPSO} algorithms achieve better result than other algorithms. The two swarm algorithms were tested with and without the ellipse fitting module. Combining the ellipse detection module improves the performance in the tested dataset.

@@

\section{Conclusion and Future Work}
\label{ConclusionandFutureWork}
This paper presented a new approach to sketch recognition using DPSO. Results show that the \textit{DPSO} in general generates an optimized stroke segmentation which improves the final recognition rate.  The final recognition is based on geometrical,statistical and structural features. The experiments evaluated two \textit{DPSO} different segmentation algorithms (AlgS1 and AlgS2) on simple presentation set dataset. Results show that AlgS2 achieves better segmentation and final accuracy than Alg3 \cite{earlyprocess} on the dataset used. The results proved that the system is easily expandable to more and more symbols as it does not depend on low level primitive recognizers but on high level symbol and geometrical features.  The system achieved an average overall improvement of more than 2\% over Alg3.  
%This paper presented a new approach to sketch recognition using PSO. The system uses both speed and curvature information which help improving the \textit{DPSO}. It is noted that the \textit{DPSO} in general generates an optimized stroke segmentation which improves the final recognition rate.  The tradeoff between accuracy achieved and time complexity must be further investigated to achieve better results. The use of statistical moments and structural features improves the recognition rate. The system was tested on 13 different symbols and achieved an accuracy of 91.5\%. The system does not depend on low level recognizer but rather on a set of high level features. This makes the system easily expandable to more and more symbols.
